---
title: "Breeding data and relationship matrices"
author: "Muhammad Farooqi"
package: BreedingExperiment
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Breeding data and relationship matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(BreedingExperiment)
```

A quantitative-genetic analysis needs four things at once: the marker
genotypes, the phenotypes, the pedigree, and where the markers sit on the
genome. Keeping them in four separate objects is where mistakes creep in, as
soon as one of them is filtered or reordered.

`BreedingExperiment` keeps them together. It extends
`RangedSummarizedExperiment`, so markers are rows, individuals are columns, and
the marker positions are a `GRanges`. Everything you already know about
Bioconductor containers applies, and subsetting can never desynchronise the
parts.

## The demonstration data

The package ships with `demoBreeding`, a simulated five-generation population
built to behave like a real programme rather than a textbook example.

```{r demo-data}
data(demoBreeding)
be <- demoBreeding
be
```

Read the last line carefully: the pedigree describes **180 individuals but only
120 were genotyped**. The two earliest generations were born before genotyping
began. That is the normal situation in a breeding programme, and it is the
reason `Hmatrix()` exists.

The data also carry the other awkward features of real records — a trait
measured on one sex only, a little missing genotype data, and unequal family
sizes because a few sires were used heavily.

```{r demo-features}
table(SummarizedExperiment::colData(be)$generation)
summary(SummarizedExperiment::colData(be)$yield)   # females only
```

If you would rather generate your own population, `simulateBreeding()` does
that.

The standard accessors work:

```{r accessors}
dim(be)
genotypes(be)[1:4, 1:5]
head(SummarizedExperiment::colData(be), 3)
head(pedigree(be), 3)
```

Because markers are genomic ranges, a data set can be cut down to a region:

```{r subset-region}
chr1 <- be[as.character(GenomicRanges::seqnames(
  SummarizedExperiment::rowRanges(be))) == "chr1", ]
dim(chr1)
```

Your own data goes in the same way:

```{r your-data , eval = FALSE}
be <- BreedingExperiment(
  genotypes  = dosage_matrix,          # markers x individuals, 0/1/2
  phenotypes = phenotype_data_frame,   # one row per individual
  pedigree   = pedigree_data_frame,    # id, sire, dam
  rowRanges  = marker_granges)         # optional, but worth having
)
```

## Quality control

```{r qc}
nrow(be)
be <- filterMarkers(be, min_maf = 0.05, min_call_rate = 0.9)
be <- imputeMarkers(be)
nrow(be)
head(alleleFrequency(be), 3)
```

## Relationships from the pedigree

`Amatrix()` builds the numerator relationship matrix. Its entries are the
familiar textbook values: one half between parent and offspring, one quarter
between grandparent and grandchild.

```{r amatrix}
A <- Amatrix(be)
round(A[1:5, 1:5], 3)
```

The diagonal carries inbreeding:

```{r inbreeding}
summary(inbreeding(be))
```

## Relationships from the markers

`Gmatrix()` is the genomic relationship matrix. It measures what actually
happened rather than what was expected: two full sibs have a pedigree
relationship of exactly 0.5, but their genomic relationship varies around it,
because Mendelian sampling gave them different halves of their parents.

```{r gmatrix}
G <- Gmatrix(be)
round(G[1:5, 1:5], 3)
```

The two agree in the aggregate, which is a useful sanity check on any data set:

```{r ag-agreement}
gen <- colnames(be)
A22 <- A[gen, gen]
cor(A22[upper.tri(A22)], G[upper.tri(G)])
```

`Dmatrix()` gives the dominance relationships, orthogonal to the additive ones,
for models that separate the two.

## Combining both: the single-step matrix

Using only `G` discards every ungenotyped relative. Using only `A` discards the
markers. `Hmatrix()` keeps both, and covers every individual in the pedigree:

```{r hmatrix}
H <- Hmatrix(be, blend = 0.05)
dim(H)
length(attr(H, "genotyped"))
```

The genotyped block comes from the markers, the ungenotyped block from the
pedigree corrected by what the markers say about genotyped relatives, and the
off-diagonal blocks connect them. Two checks confirm the behaviour: blending
entirely towards the pedigree returns `A22`, and the result stays positive
definite.

```{r hmatrix-checks}
H1 <- Hmatrix(be, blend = 1, tune = FALSE)
max(abs(H1[gen, gen] - A22))
min(eigen(H, only.values = TRUE)$values) > 0
```

`H` is the relationship structure used by single-step genomic prediction. Pass
it, or its inverse, to whichever mixed-model solver you prefer.

## Session information

```{r session}
sessionInfo()
```
